### 2020-09-01 paper/sex_selection_spacing-ver4.tex

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

% Demography abstract
\noindent 
In India, the most substantial lengthening in Hindu women's birth intervals over the past 
four decades has occurred for the women most likely to use sex-selective abortions. 
For well-educated women with two girls, the median birth interval before the third child 
has increased by nearly 15 months---and the 75th percentile interval by 21 months---as the 
sex ratio has approached 80\% boys. 
Because of the rapid lengthening in spacing, standard fertility rates have significantly 
overestimated how fast cohort fertility fell. 
Despite convergence, cohort fertility is still 10\%--20\% higher than the 
fertility rate and above replacement level for all but the best-educated urban women. 
With the lengthening from sex selection, some women show a reversal of the traditional 
spacing pattern; instead of the shortest birth interval in the absence of sons, they now 
have the longest. 
Finally, infant mortality has declined over time for all groups, but the decline was 
fastest for the less educated. 
Short birth spacing is still associated with higher mortality, although considerably less 
so for the best-educated women. 
There is no evidence that the increasing use of sex selection is associated with higher 
infant mortality. 
Neither does it appear that the use of sex selection is declining.

@@

Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?
In India, birth intervals have traditionally been shorter with fewer sons, contributing 
to the higher mortality risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Therefore, longer birth spacing, whether from sex selection or secular changes, may reduce 
mortality \citep{Conde-Agudelo2012,Molitoris2019}.
However, a counteracting effect is possible if longer birth intervals arise from multiple 
abortions because the short duration between pregnancies could increase mortality. 
Nevertheless, there has been no attempt to examine how infant mortality has changed with 
the increase in sex selection.

@@

Finally, infant mortality has declined substantially over time for all groups, but fastest 
for the less educated, who are now close to the level of the best-educated women. 
However, mortality is still inversely related to education level, especially for very short 
birth intervals. 
There is no evidence that the increasing use of sex selection is associated with higher mortality.

@@

Direct information on the use of sex selection is not available, so I compare periods 
based on the changes in access and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--1983, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
% Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Finally, although sex selection increased even after 1994, we may have passed a 
turning point in its use in the mid-2000s \citep{Das_Gupta2009,Kumar2012,Bongaarts2013,Diamond-Smith2015}.

@@

% line 3661 in an_infant_mortality.log
The raw numbers for women with the most uneven sex ratio also suggest that even with 
very high use of sex selection, there is no impact on child mortality. 
A total of 1,004 women with 12 or more years of education and no boys at the start of 
the third spell in the last period had a third child, of which 685 were boys. 
Of these 685 boys, only six died within the first year of life. 
Half of those who died were born in the 9--32-month interval, and none in the 57-month+ 
interval.

@@

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months from the prior birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that substantial use of sex selection is associated with higher 
infant mortality.

% Implications 
An important implication of the results here is that population growth in India may
not have fallen as much as recently suggested.
There has been a rapid reduction in infant mortality risk, the longer birth spacing should 
further reduce mortality risk, and cohort fertility is higher than previously realized.

@@

There are two critical questions that future research should address.
First, sex selection means that girls-only families are less likely to have
very short birth intervals, which may reduce sibling competition. 
Hence, better health outcomes for girls with sex selection could be an unintended 
side-effect, rather than the result of girls becoming more valued as is often assumed. 
Comparing prior children's outcomes across sex composition and the sex of the next child 
could be a way to understand why girls' health outcomes improve in the presence of sex selection.

>>>>

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Monica Das Gupta, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, three anonymous reviewers, and seminar participants at University of 
Copenhagen, University of Michigan, University of Washington, University of Aarhus, the 
Fourth Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

% Demography abstract
\noindent 
% In India, the most substantial lengthening in Hindu women's birth intervals over the past 
% four decades has occurred for the women most likely to use sex-selective abortions. 
% For well-educated women with two girls, the median birth interval before the third child 
% has increased by nearly 15 months---and the 75th percentile interval by 21 months---as the 
% sex ratio has approached 80\% boys. 
% Because of the rapid lengthening in spacing, standard fertility rates have significantly 
% overestimated how fast cohort fertility fell. 
% Despite convergence, cohort fertility is still 10\%--20\% higher than the 
% fertility rate and above replacement level for all but the best-educated urban women. 
% With the lengthening from sex selection, some women show a reversal of the traditional 
% spacing pattern; instead of the shortest birth interval in the absence of sons, they now 
% have the longest. 
% Finally, infant mortality has declined over time for all groups, but the decline was 
% fastest for the less educated. 
% Short birth spacing is still associated with higher mortality, although considerably less 
% so for the best-educated women. 
% There is no evidence that the increasing use of sex selection is associated with higher 
% infant mortality. 
% Neither does it appear that the use of sex selection is declining.

\noindent Over the past four decades, the Hindu women in India most likely to use 
sex-selective abortions---well-educated women with no sons---had the most substantial 
lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some
women with no sons having longer birth intervals than those with sons. 
Those least likely to use sex-selective abortions---less-educated women in rural 
areas---still follow the traditional pattern of short spacing when they have girls, with 
only limited evidence of sex selection. 
Because of the rapid lengthening in spacing, the standard fertility rates substantially 
overestimated how fast cohort fertility fell. 
Despite a convergence, cohort fertility is still 10\%--20\% higher than the fertility rate 
and above replacement level for all but the best-educated urban women. 
Infant mortality has declined substantially over time for all groups, with the fastest 
decline among the less educated. 
Short birth spacing is still associated with higher mortality, although considerably less 
so for the best-educated women. 
There is no evidence that repeated sex-selective abortions are associated with higher 
infant mortality for the child eventually born. 
Finally, it does not appear that the use of sex selection is declining.

@@

Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?
In India, birth intervals have traditionally been shorter with fewer sons, contributing 
to the higher mortality risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Therefore, longer birth spacing, whether from sex selection or secular changes, may reduce 
mortality \citep{Conde-Agudelo2012,Molitoris2019}.
However, a counteracting effect is possible if longer birth intervals arise from multiple 
abortions because the short duration between pregnancies could increase mortality. 
% Nevertheless, there has been no attempt to examine how infant mortality has changed with 
% the increase in sex selection.

@@

Finally, infant mortality has declined substantially over time for all groups, but fastest 
for the less educated, who are now close to the level of the best-educated women. 
However, mortality is still inversely related to education level, especially for very short 
birth intervals. 
There is no evidence that repeated sex-selective abortions are associated with higher 
mortality for the child eventually born.

@@

Direct information on the use of sex selection is not available, so I compare periods 
based on the changes in access and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--1983, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
% Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2019}.
}
Finally, although sex selection increased even after 1994, we may have passed a 
turning point in its use in the mid-2000s \citep{Das_Gupta2009,Kumar2012,Bongaarts2013,Diamond-Smith2015}.

@@

% line 3661 in an_infant_mortality.log
The raw numbers for women with the most uneven sex ratio also suggest that even with 
very high use of sex selection, there is no impact on mortality. 
A total of 1,004 women with 12 or more years of education and no boys at the start of 
the third spell in the last period had a third child, of which 685 were boys. 
Of these 685 boys, only six died within the first year of life. 
Half of those who died were born in the 9--32-month interval, and none in the 57-month+ 
interval.

@@

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months from the prior birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that repeated abortions are associated with higher infant mortality 
for the child eventually born.

% Implications 
The results here paint a less rosy picture of India's prospects for a continued reduction 
in population growth than generally accepted. 
With predicted cohort fertility still substantially higher than the fertility rate, 
India's total fertility rate will likely stabilize or even increase as birth intervals slow 
their lengthening. 
The more successful the attempts at combatting sex selection are, the more likely an 
increase in the total fertility rate will be. 
Furthermore, the rapid decline in infant mortality risk, combined with likely future 
declines as the proportion of very short birth intervals falls, may also slow the 
reduction in population growth.

@@

There are two critical questions that future research should address.
First, sex selection means that girls-only families are less likely to have
very short birth intervals, which may reduce sibling competition. 
Hence, better health outcomes for girls with sex selection could be an unintended 
side-effect, rather than the result of girls becoming more valued as is often 
assumed \citep{Hu2015}. 
Comparing prior children's outcomes across sex composition and the sex of the next child 
could be a way to understand why girls' health outcomes improve in the presence of sex selection.
